Monte Carlo estimation of the probability of causal contacts between communicating civilisations

### 2020-06-04 submit_v03/manuscript.tex
submitted V3

\begin{figure}[h]
%
\centering
%
\includegraphics[width=\columnwidth]{F_example.pdf}
%
\caption{Example of a configuration of an emitter and a receiver 
   to produce a causal contact.
   %
   The observers are separated by 100~light years and appear with a
   difference in time of 80~years. One observer is active between
   $t=$0~yr and $t=$50~yr.  The other observer is also active for
   50~yr, between $t=$80~yr and $t=$130~yr. 
   %
   The causal connection occurs between $t=$100~yr and $t=$130~yr
   %
   Meaningfull times are marked with vertical dashed lines.
   }
%
\label{F_example}
%
\end{figure}

@@

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node.
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centred on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
This sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The node E$_1$ can listen signals from the node E$_2$, since
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the node E$_2$ is in causal contact from the
   node E$_1$, in the time interval (C$_{21}$, B$_{21}$).
%
} \label{F_scheme}
%
\end{figure}

@@

In what follows, we outline the experimental setup adopted to estimate the
probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the mean time
span between the appearance of consecutive \cetis{}, $\left<\tau_{\rm
a}\right>$,
their mean lifetime $\left<\tau_{\rm s}\right>$, and the maximum distance a signal
can be detected by another \ceti{} ($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts between pairs
of \cetis{} would be larger for smaller $\tau_{\rm a}$ parameters,
higher $\tau_{\rm s}$ or
higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans
($\tau_{\rm s}$) and the number of \cetis{} per unit time
\citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive
\cetis{} (since when $\left<\tau_{\rm a}\right>$ is shorter, it produces a greater
density of \cetis{}).
%
The analytical expresions for these distributions are set to a fixed law, as
discussed in Sec.~\ref{SS_PDF_shape}.

\ffn{1} Space--time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Fig.~\ref{F_scheme}.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon.
%
This polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

In our experimental configuration the simulation starts assuming that the
stochastic process is already stable, and finishes before any galactic
evolution effect could modify the fixed values of the variables.
%
Likewise, we assume that the probability for the appearance of a \ceti{} is
homogeneous over the GHZ.
%
Although the Galaxy has a well-known spiral structure, the nodes are assumed
to be sparse (otherwise the Galaxy would be full of life) and the spiral
structure would not, in that case, produce significant differences.
%
If the distribution of nodes is not sparse, as it could be the case if the
spiral arms host most of the \cetis{}, then contacts would be more frequent
between closely located nodes.
%
In such a case our results underestimate the number of contacts between
close pairs of nodes within the same spiral arm, and conversely, overestimate
the number of contacts between separated nodes.
%
We also limit the possibilities of life or other types of civilisations to the
usually stated hypotheses for the definition of the GHZ
\citep{dayal_habitability_2016, gonzalez_galactic_2001,
lineweaver_galactic_2004, gonzalez_habitable_2005, morrison_extending_2015,
haqq-misra_evolution_2019, rahvar_cosmic_2016, gobat_evolution_2016,
rahvar_cosmic_2016} and consider that habitability remains constant over time
\citep[see, however, ][]{gonzalez_habitable_2005, dayal_habitability_2016,
gobat_evolution_2016}.
%
This means that we set aside civilisations that could survive in severe
conditions or unstable systems, which would prevent the appearance of life as
we know it.

@@

In our simulations, we assume the simplest configuration for the
growth of the sphere of first contact.
%
In particular, we do not consider the
possibility of stellar colonisation
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000,
maccone_mathematical_2011}.
%
We also assume that communication is
equally likely in all directions, i.e., we assume isotropic
communication in all cases.
%
Different communication efficiencies or
detection methods are straightforward to carry out in the simulations
for more detailed and complex scenarios.
%
This approach, however, is
beyond this work because it obscures the experimental setup and make
the results less clear.
%
The assumptions we accept imply that the
results are independent of whether intelligent agents are organic or
artificial.
%
Moreover, the causes of the limited lifetime of a
civilisation can be natural (astrophysical phenomena), caused by auto
destruction or by external factors, to name a few.
%
However, we assume
that these events are sufficiently numerous in the Galaxy so that a
statistical model is plausible. The failure of this hypothesis would
imply extreme values for the parameters that represent the density of
the nodes (i.e., $\left< \tau_{\rm a} \right>$).

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (parameter space)}
   &min. value&max. value&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings 
	      & 5$\cdot10^2$~yr & 40500~yr & 21\\ 
	    & & 5$\cdot$10$^5$~yr & 10$^6$~yr & 51\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
	      & 5$\cdot10^2$~yr & 40500~yr & 21\\ 
	    & & 5$\cdot$10$^5$~yr & 10$^6$~yr & 51\\ 
	D$_{\rm max}$ & Maximum reach of a message  & \multicolumn{3}{c}{100 pc, 500 pc, 1 kpc, 5 kpc, 10 kpc} \\
\hline
   \multicolumn{2}{l}{fixed variables} & \multicolumn{2}{c}{assumptions} &value \\
\hline
   & statistical properties of all \cetis{} &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & two-dimensional ring &&\\
	R$_{GHZ}^{\rm min}$   & Inner radius of the GHZ  & \citet{lineweaver_galactic_2004} & & 7~kpc\\
   R$_{GHZ}^{\rm max}$   & Outer radius of the GHZ       & \citet{lineweaver_galactic_2004} & & 9~kpc\\
	t$_{\rm max}$ & Time span of the simulation  & &
   \multicolumn{2}{r}{10$^5$~yr - 10$^9$~yr} \\
    & \multicolumn{3}{l}{number of random realizations for each point
    in the parameter space} & 1 - 50 \\
\hline
   \multicolumn{2}{l}{discrete events} &
   \multicolumn{3}{l}{affected variables}\\
\hline
   A event & Awakening: a \ceti{} starts its communication
   capabilities &Number of active \cetis{}\\
   B event & Blackout: the end of the communication channel stops
   &Number causal channels\\
   C event & Contact: a new causal contact is produced &Number of causal channels\\
   D event & Doomsday: a \ceti{} ends its communication capabilities&Number of active \cetis{}\\
\hline

@@

The power law and exponential statistical distributions are among the most
common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many languages is
known to follow the law of Zipf (which is a power law).
%
These distributions arise any time a phenomenon is characterised by commonly
occurring small events and rarely produced large events \citep[e.g.
][]{adamic_zipf_2000}.
%
Zipf law also describes population ranks of cities in various countries,
corporation sizes, income rankings, ranks of the number of people watching the
same TV channel, etc.
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and floods; the
sizes of meteorites or the losses caused by business interruptions from
accidents, are also well described by power laws
\citep{sornette_critical_2006}.
%
%% , benguigui_classificacion_2016}.
%The power law behaviour has been observed in a variety of systems.
%
Additional examples include stock market fluctuations, sizes of computer files
or word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}. 
%
Power laws have also been widely used in biological sciences.
%
Some examples are the analysis of connectivity patterns in metabolic networks
\citep{jeong_large_2000} or the number of species observed per unit area in
ecology \citep{martin_origin_2006, frank_common_2009}.
%
More examples can be found in the literature \citep{martin_origin_2006,
maccone_KLT_2010} \citep{barabasi_scale_2009, maccone_evolution_2014}
\citep{maccone_lognormals_2014}
%
This distribution family is suitable for the statistical description
of the $D_{\rm mar}$ parameter, although in this work we assume a uniform
distribution for simplicity.
%
The power law family of functions is also a good candidate for the description of
the temporal variables in the model.
%
However, we prefer the exponential model.
%
The exponential distribution of lifespan and waiting times is justified by
considering the hypothesis that the process of appearance of life in the galaxy
is homogeneous and stationary.
%
That is, there is not a preferred region within
the GHZ for the spontaneous appearance of life, and the emergence of a node is
independent of the existence of previous nodes in the galaxy.
%
These seem to be simple conditions, and allow to propose a distribution family
without knowing the details of the underlying process.
%
The exponential distribution for the separations in time is equivalent to
proposing a Poisson process for the emergence of nodes, given the relation
between the number of events in time or space and the waiting time or
separation, respectively \citep[e.g., ][]{ross_simulation_2012}.
%
That is, these
are two alternative approaches to describing the same process, a Poisson
distribution for the number of events implies an exponential distribution for
their separations, and vice versa.
%
It should be emphasised that the exponential
laws used in this work are assumed as part of the working hypothesis, and
instead of analysing results from a particular parameter chosen ad hoc, we
explore the parameter space and analyse the impact of the values of these
parameters on the results.

\begin{figure} \centering
   \includegraphics[width=0.5\textwidth]{F_number_of_contacts.pdf}
   \caption{
   %
Empirical cumulative distributions of the number contacts
   for \cetis{} in different samples and with D$_{\rm max}$=10~kpc.
%
The upper panel shows the variation of the distribution
   as a function of the $\tau_{\rm a}$ parameter, including the values 
   0.1, 0.5, 1, 2, and 5~kyr. All the curves
   correspond to models with 
   $\tau_{\rm s}=$40~kyr and \mbox{D$_{\rm max}=10$~kpc}.
%
   The bottom panel corresponds to values of $\tau_{\rm s}$
   5, 10, 20 and 50~kyr,
   $\tau_{\rm a}=1$~kyr and D$_{\rm max}=10$~kpc.
%
   Cumulative distributions can be used to visualise the probability
   of a random node of having more than M contacts in a given model.  For
   example, in the upper panel the probability of a random node having
   more than 10 contacts is nearly one for the model with the
   smallest $\tau_{\rm a}$ value, and nearly zero for the model with the
   largest $\tau_{\rm a}$ value.
   %
   } \label{F_number_of_contacts}
\end{figure}

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_never_contact.pdf}
   \caption{
   %
Rate of nodes that never make contact (listening) as a
   function of $\tau_{\rm a}$ and $\tau_{\rm s}$, for 
$D_{\rm max}$=1~kpc (left panel),
$D_{\rm max}$=5~kpc (middle panel), and
$D_{\rm max}$=10~kpc (right panel)
%
   }
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_C_at_A.pdf}
   \caption{
   %
Rate of nodes that listen at the moment of the
   awakening, as a funcion of $\tau_{\rm a}$ and $\tau_{\rm s}$, for
$D_{\rm max}$=1~kpc (left panel),
$D_{\rm max}$=5~kpc (middle panel), and
$D_{\rm max}$=10~kpc (right panel)
%
   }
   \label{F_C_at_A}
\end{figure*}

\subsection{Model for \cetis{}}

Since the method is based on events, the first step is to define the
relationships between them and how events trigger changes on the state of the
system.
%
Once an event is produced, the system is updated by saving the corresponding
change in the related variables.
%
For example, a B-type event diminishes the number of active \cetis{} by one,
and also could affect the number of ongoing communications.
%
We consider a first approach to the galactic habitable zone as a two-dimensional
annular region.
%
The adopted values for the geometry of the GHZ are 7~kpc and 9~kpc, for the
inner and outer radius, respectively \citep{lineweaver_galactic_2004}. 
%
This simple model does not take into account the variations in stellar density
given by the spiral structure of the Galaxy.
%
The main task in setting a DE simulation is to define the events and what
changes do these events produce on the system.
%
Although there are several possible approaches, we choose to follow the
evolution of the system according to the following events:

\begin{enumerate}
   \item[(A)] Awakening: A new \ceti{} acquires the ability to emit
      and receive messages and starts to actively broadcast and
      lookout for messages.
   \item[(C)] Contact: A new one-way causal contact is established.
   \item[(B)] Blackout: An existing causal contact is interrupted,
      with or without the occurrence of a doomsday event.
   \item[(D)] Doomsday: An old \ceti{} disappears, or halts its
      communication activities.
\end{enumerate}

We have chosen this nomenclature so that it can be easily remembered
from the initials A, B, C, D.
%
The \blackout is produced when one of the two \cetis{} halts in its
capability of emitting and receiving signals.
%
It can occur at the same time the \doomsday or before.
%
Similarly, a \ccontact can be produced at the time of the \aawakening
or later, and there can be several \contacts for the same node.
%
The time marks for each event (A and D for each \ceti{} and C and B
for each constrained causal contact) are stored as a result of the simulation.
%
Also, the list of active \cetis{} is obtained as a function of time.

\subsection{Model implementation}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{F_waiting_for_1C.pdf}
   \caption{
   %
Histograms of the mean waiting times for the first contact, for
several models.
%
   Left panel shows the histograms for several values of $\tau_{\rm a}$, and
   $\tau_{\rm s}$ in the range 17000-24000~yr.
%
   Right panel shows the histograms for several values of $\tau_{\rm
   s}$, and
   $\tau_{\rm a}$ in the range 68000-100000~yr.
%
   The shape of the histograms does not change when $\tau_{\rm a}$ is
   modified, but significantly changes when $\tau_{\rm s}$ is modified. 
%
   }
   \label{F_waiting_for_1C}
\end{figure}

>>>>

\begin{figure}[h]
%
\centering
%
\includegraphics[width=\columnwidth]{F_example.pdf}
%
\caption{Example of a configuration of an emitter and a receiver 
   to produce a causal contact.
   %
   The observers are separated by 100~light years and appear with a
   difference in time of 80~years. One observer is active between
   $t=$0~yr and $t=$50~yr.  The other observer is also active for
   50~yr, between $t=$80~yr and $t=$130~yr. 
   %
   The causal connection occurs between $t=$100~yr and $t=$130~yr.
   %
   Meaningful times are marked with vertical dashed lines.
   }
%
\label{F_example}
%
\end{figure}

@@

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node.
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of the first contact (SFC) is the sphere centred on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
This sphere is represented by the left triangle of the region
$R$.
%
The surface of the last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The node E$_1$ can listen to signals from the node E$_2$, since
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the node E$_2$ is in causal contact from the
   node E$_1$, in the time interval (C$_{21}$, B$_{21}$).
%
} \label{F_scheme}
%
\end{figure}

@@

In what follows, we outline the experimental setup adopted to estimate the
probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the mean time
span between the appearance of consecutive \cetis{}, $\tau_{\rm a}$,
their mean lifetime $\tau_{\rm s}$, and the maximum distance a signal
can be detected by another \ceti{} ($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts between pairs
of \cetis{} would be larger for smaller $\tau_{\rm a}$ parameters,
higher $\tau_{\rm s}$ or
higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans
($\tau_{\rm s}$) and the number of \cetis{} per unit time
\citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive
\cetis{} (since when $\tau_{\rm a}$ is shorter, it produces a greater
density of \cetis{}).
%
The analytical expresions for these distributions are set to a fixed law, as
discussed in Sec.~\ref{SS_PDF_shape}.

\ffn{1} Space--time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Fig.~\ref{F_scheme}.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon.
%
This polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$.
%
The receiver node E$_1$ can listen to signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen to signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

In our experimental configuration the simulation starts assuming that the
stochastic process is already stable, and finishes before any galactic
evolution effect could modify the fixed values of the variables.
%
Likewise, we assume that the probability for the appearance of a \ceti{} is
homogeneous over the GHZ.
%
The adopted geometry of the GHZ in all the simulations is given by a 
two-dimensional annular region,
with an inner radius of 7~kpc and an outer radius of 9~kpc
\citep{lineweaver_galactic_2004}.
%
Although the Galaxy has a well-known spiral structure, the nodes are assumed
to be sparse (otherwise the Galaxy would be full of life) and the spiral
structure would not, in that case, produce significant differences.
%
If the distribution of nodes is not sparse, as it could be the case if the
spiral arms host most of the \cetis{}, then contacts would be more frequent
between closely located nodes.
%
In such a case our results underestimate the number of contacts between
close pairs of nodes within the same spiral arm, and conversely, overestimate
the number of contacts between separated nodes.
%
We also limit the possibilities of life or other types of civilisations to the
usually stated hypotheses for the definition of the GHZ
\citep{dayal_habitability_2016, gonzalez_galactic_2001,
lineweaver_galactic_2004, gonzalez_habitable_2005, morrison_extending_2015,
haqq-misra_evolution_2019, rahvar_cosmic_2016, gobat_evolution_2016,
rahvar_cosmic_2016} and consider that habitability remains constant over time
\citep[see, however, ][]{gonzalez_habitable_2005, dayal_habitability_2016,
gobat_evolution_2016}.
%
This means that we set aside civilisations that could survive in severe
conditions or unstable systems, which would prevent the appearance of life as
we know it.

@@

In our simulations, we assume the simplest configuration for the
growth of the sphere of first contact.
%
In particular, we do not consider the
possibility of stellar colonisation
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000,
maccone_mathematical_2011}.
%
We also assume that communication is
equally likely in all directions, i.e., we assume isotropic
communication in all cases.
%
Different communication efficiencies or
detection methods are straightforward to carry out in the simulations
for more detailed and complex scenarios.
%
This approach, however, is
beyond this work because it obscures the experimental setup and make
the results less clear.
%
The assumptions we accept imply that the
results are independent of whether intelligent agents are organic or
artificial.
%
Moreover, the causes of the limited lifetime of a
civilisation can be natural (astrophysical phenomena), caused by auto
destruction or by external factors, to name a few.
%
However, we assume
that these events are sufficiently numerous in the Galaxy so that a
statistical model is plausible. The failure of this hypothesis would
imply extreme values for the parameters that represent the density of
the nodes (i.e., $\tau_{\rm a}$).

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (parameter space)}
   &min. value&max. value&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings 
	      & 5$\cdot10^2$~yr & 40500~yr & 21\\ 
	    & (linear grids) & 5$\cdot$10$^5$~yr & 10$^6$~yr & 51\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
	      & 5$\cdot10^2$~yr & 40500~yr & 21\\ 
	    & (linear grids) & 5$\cdot$10$^5$~yr & 10$^6$~yr & 51\\ 
	D$_{\rm max}$ & Maximum reach of a message  & \multicolumn{3}{c}{100 pc, 500 pc, 1 kpc, 5 kpc, 10 kpc} \\
\hline
   \multicolumn{2}{l}{fixed variables} & \multicolumn{2}{c}{assumptions} &value \\
\hline
   & statistical properties of all \cetis{} &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & two-dimensional ring &&\\
	R$_{GHZ}^{\rm min}$   & Inner radius of the GHZ  & \citet{lineweaver_galactic_2004} & & 7~kpc\\
   R$_{GHZ}^{\rm max}$   & Outer radius of the GHZ       & \citet{lineweaver_galactic_2004} & & 9~kpc\\
	t$_{\rm max}$ & Time span of the simulation  & &
   \multicolumn{2}{r}{10$^5$~yr - 10$^9$~yr} \\
    & \multicolumn{3}{l}{number of random realizations for each point
    in the parameter space} & 1 - 50 \\
\hline
   \multicolumn{2}{l}{discrete events} &
   \multicolumn{3}{l}{affected variables}\\
\hline
   A event & Awakening: a \ceti{} starts its communication
   capabilities &Number of active \cetis{}\\
   B event & Blackout: the end of the communication channel stops
   &Number causal channels\\
   C event & Contact: a new causal contact is produced &Number of causal channels\\
   D event & Doomsday: a \ceti{} ends its communication capabilities&Number of active \cetis{}\\
\hline

@@

The power law and exponential statistical distributions are among the most
common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many languages is
known to follow the law of Zipf (which is a power law).
%
These distributions arise any time a phenomenon is characterised by commonly
occurring small events and rarely produced large events \citep[e.g.
][]{adamic_zipf_2000}.
%
Zipf law also describes population ranks of cities in various countries,
corporation sizes, income rankings, ranks of the number of people watching the
same TV channel, etc.
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and floods; the
sizes of meteorites or the losses caused by business interruptions from
accidents, are also well described by power laws
\citep{sornette_critical_2006}.
%
Additional examples include stock market fluctuations, sizes of computer files
or word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}. 
%
Power laws have also been widely used in biological sciences.
%
Some examples are the analysis of connectivity patterns in metabolic networks
\citep{jeong_large_2000} or the number of species observed per unit area in
ecology \citep{martin_origin_2006, frank_common_2009}.
%
More examples can be found in the literature \citep{martin_origin_2006,
maccone_KLT_2010, barabasi_scale_2009, maccone_evolution_2014, maccone_lognormals_2014}.
%
This distribution family is suitable for the statistical description
of the $D_{\rm max}$ parameter, although in this work we assume a uniform
distribution for simplicity.
%
The power law family of functions is also a good candidate for the description of
the temporal variables in the model.
%
However, we prefer the exponential model.
%
The exponential distribution of lifespan and waiting times is justified by
considering the hypothesis that the process of appearance of life in the galaxy
is homogeneous and stationary.
%
That is, there is not a preferred region within
the GHZ for the spontaneous appearance of life, and the emergence of a node is
independent of the existence of previous nodes in the galaxy.
%
These seem to be simple conditions, and allow to propose a distribution family
without knowing the details of the underlying process.
%
The exponential distribution for the separations in time is equivalent to
proposing a Poisson process for the emergence of nodes, given the relation
between the number of events in time or space and the waiting time or
separation, respectively \citep[e.g., ][]{ross_simulation_2012}.
%
That is, these
are two alternative approaches to describing the same process, a Poisson
distribution for the number of events implies an exponential distribution for
their separations, and vice versa.
%
It should be emphasised that the exponential
laws used in this work are assumed as part of the working hypothesis, and
instead of analysing results from a particular parameter chosen ad hoc, we
explore the parameter space and analyse the impact of the values of these
parameters on the results.

\begin{figure} \centering
   \includegraphics[width=0.5\textwidth]{F_number_of_contacts.pdf}
   \caption{
   %
Empirical cumulative distributions of the number contacts
   for \cetis{} in different samples and with D$_{\rm max}$=10~kpc.
%
The upper panel shows the variation of the distribution
   as a function of the $\tau_{\rm a}$ parameter, including the values 
   0.1, 0.5, 1, 2, and 5~kyr. All the curves
   correspond to models with 
   $\tau_{\rm s}=$40~kyr and \mbox{D$_{\rm max}=10$~kpc}.
%
   The bottom panel corresponds to 
   $\tau_{\rm a}=1$~kyr and D$_{\rm max}=10$~kpc, and the following values
   of $\tau_{\rm s}$: 5, 10, 20 and 50~kyr.
%
   Cumulative distributions can be used to visualise the probability
   of a random node of having more than M contacts in a given model.  For
   example, in the upper panel the probability of a random node having
   more than 10 contacts is nearly one for the model with the
   smallest $\tau_{\rm a}$ value, and nearly zero for the model with the
   largest $\tau_{\rm a}$ value.
   %
   } \label{F_number_of_contacts}
\end{figure}

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_never_contact.pdf}
   \caption{
   %
The fraction of constrained causal contact nodes that never make contact (listening) as a
   function of $\tau_{\rm a}$ and $\tau_{\rm s}$, for 
$D_{\rm max}$=1~kpc (left panel),
$D_{\rm max}$=5~kpc (middle panel), and
$D_{\rm max}$=10~kpc (right panel).
%
The values of $\tau_{\rm a}$ and $\tau_{\rm s}$ are in the range
   5$\cdot10^5$ to 10$^6$~yr.
%
The matrix is shown as obtained from the simulations, and the level
   curves are shown for the smoothed matrix.
%
   }
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_C_at_A.pdf}
   \caption{
   %
The fraction of constrained causal contact nodes that make contact at the moment of the
   awakening (i.e., $t_{\rm A}=t_{\rm C}$), as a function of $\tau_{\rm a}$ and $\tau_{\rm s}$, for
$D_{\rm max}$=1~kpc (left panel),
$D_{\rm max}$=5~kpc (middle panel) and
$D_{\rm max}$=10~kpc (right panel).
%
The values of $\tau_{\rm a}$ and $\tau_{\rm s}$ are in the range
   5$\cdot10^5$ to 10$^6$~yr.
%
The matrix is shown as obtained from the simulations, and the level
   curves are shown for the smoothed matrix.
   }
   \label{F_C_at_A}
\end{figure*}

\subsection{Implementation of the SC3Net model}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{F_waiting_for_1C.pdf}
   \caption{
   %
Histograms of the mean waiting times for the first contact, for
several models.
%
   Upper panel shows the histograms for several values of $\tau_{\rm a}$, and
   $\tau_{\rm s}$=10~kyr and D$_{\rm max}$=10~kpc.
%
   Bottom panel shows the histograms for several values of $\tau_{\rm
   s}$, with $\tau_{\rm a}$=10~kyr and D$_{\rm max}$=10~kpc. The
   arrows indicate the sample and highlight the values of the first
   waiting time bin, which has a clear excess for the models with
   larger $\tau_{\rm s}$ values.
%
   The choice of these models was made in order to show the trends in
   the results as a function of the two temporal parameters.
%
   }
   \label{F_waiting_for_1C}
\end{figure}
